Testing Gravity using Type Ia Supernovae Discovered by LSST

### 2018-12-14 doc/article.tex
A version, not quite smooth, to give to Eric.

The precision in  $f\sigma_8$ derived from the baseline WFD ten-year SN~Ia discoveries has been projected by \citet{2017ApJ...847..128H},
from both their RSD and peculiar velocities.
The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance uncertainty is assumed for each supernova.
SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy, cluster, and Ly$\alpha$
RSD measurements are sensitive.  A joint galaxy RSD and SN peculiar velocity will be even more powerful.
{\bf Add a SN PV + DESI-like BGS column and/or a PV-only column with Cullan's code? }

% Requires the booktabs if the memoir class is not being used
\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{@{} lccc @{}} % Column formatting, @{} suppresses leading/trailing space
	\hline
	Redshift & RSD & RSD + PV & cumulative\\ \hline
      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\
      \hline
   \end{tabular}
   \caption{Projected percent uncertainties in $f\sigma_8$ from a 10-year LSST SN survey with 5\% distance uncertainties from
   \citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
   The cumulative column shows the effective uncertainty from combining the redshift and all other shallower redshift bins.}
   \label{tab:howlett}
\end{table}

@@

\section{Designing a Supernova Survey to Measure $\MakeLowercase{f}\sigma_8$}
\subsection{Parameters that Determine Scientific Performance}

\citet{2017ApJ...847..128H} perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the covariance for the velocity-velocity correlation
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\label{cov:eq}
\end{equation}
is dependent on the power spectrum, noise in the velocity measurement dominated by the intrinsic velocity dispersion $\sigma$, and the density of velocity probes
\citep{2017MNRAS.464.2517H}.  
The primary dependence on the growth of structure is through
\begin{equation}
\frac{\partial C}{\partial \lambda} = \frac{\partial P_{vv}}{\partial \lambda} = \text{constant}
\end{equation}
for the parameter choice
$\lambda=(f\sigma_8)^2$.  In the sample-variance limit, $F_{\lambda \lambda} \propto V$ whereas in the shot noise limit $F_{\lambda \lambda} \propto V n^2 \sigma^{-4}$.

The survey parameters that determine parameter precision are the volume $V$, which
in turn can be parameterized by the survey solid angle $\Omega$ and the redshift depth $z_{max}$ ; the number density of sources $n$; and the intrinsic
magnitude dispersion $\sigma_M$, where magnitude and velocity dispersions are related by $\sigma_{M} = \frac{5}{\ln{10}} \frac{1+z}{z} \sigma$.

LSST is expected to discover all $z<0.3$ SNe~Ia before maximum light in its active Wide Fast Deep (WFD) survey area.
Equating LSST discovery with having a meaningful peculiar velocity, the number density is$n$ is thus a non-tunable parameter, which only increases with survey duration.
The remaining free parameters are then $\Omega$, $z_{max}$, and $\sigma_M$.

While LSST data can be used to obtain photometric redshifts from host galaxies and supernova light curves, the spectroscopic
redshifts needed for precise host-galaxy peculiar velocities require other resources.

\begin{itemize}
\item Solid Angle $\Omega$:
The variance in $f\sigma_8$ is proportional to solid angle $\Omega$.  The baseline
LSST survey covers 18,000 sq.~deg, at $-75 \lesssim \delta \lesssim15$ avoiding the Galactic plane.  Larger solid-angle coverage for $z<0.3$ SN~Ia searches, beyond
the LSST baseline, benefits peculiar-velocity science.  Complementary northern-hemisphere surveys and LSST-expanded or independent
coverage of the southern equatorial pole, could double
the sky coverage and halve the variance in $f\sigma_8$.
\item Redshift Depth $z_{max}$:
For the number densities and intrinsic magnitude dispersions relevant to LSST, the noise is not sample-variance
limited.  We have run simulations of LSST SN surveys
and calculated the signal-to-noise (STON) of $f\sigma_8$ from peculiar velocities. These simulations show, as seen in the left plot
of  Figure~\ref{notshot:fig}, that an increase in the number the number density of tracers leads to significant improvements in STON.
It is also worth noting that sample variance is not negligible, as seen by the non-linear dependence of  STON on $\sigma_M^2/n$.

@@

The effective contribution to the STON per supernova, quantified as STON divided by the root of the number of tracers as shown in the
right plot of   Figure~\ref{notshot:fig}, shows that LSST is in the regime where adding one tracer at lower redshift better improves the 
STON relative to the addition of one at higher redshift.

@@

\item Intrinsic Magnitude Dispersion $\sigma_M$:
Just as an increase in tracer number density beyond that of a ten-year LSST survey improves the STON in the measurement
of $f\sigma_8$, as noted in the discussion on $z_{max}$, so too does a decrease in the supernova intrinsic magnitude dispersion.

The nominal LSST WFD observing plan is to have
each piece of sky observed every couple of days rotating through the 6 $grizy$ bands. The data is sufficient to have effectively
complete $z<0.3$ SN~Ia discoveries.  However,  the per-band light curves are poorly sampled, which limits the accuracy with which distances
can be derived.
LSST SN~Ia light curves from the WFD survey, with their relatively poor sampling, plus a spectroscopic redshift
 can give $\sigma_M \sim 0.15$~mag.

Supplemental non-LSST follow-up data in the form or improved temporal light-curve sampling (accurate rise and decline times), 
expanded (UV, NIR) wavelength coverage, and spectral features
can access high-fidelity SN~Ia models with lower intrinsic magnitude dispersion and residual systematic bias.
For example, infrared data \citep{2012MNRAS.425.1007B} or spectrophotometry at peak brightness
 \citep{2015ApJ...815...58F} are projected to give
$\sigma_M \lesssim 0.08$~mag.
Uncertainties in $f\sigma_8$ depend on $\sigma^2_M/n$ so the improvement between 0.08 and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or in survey duration. 
\end{itemize}

\subsection{Designing an SN~Ia Peculiar Velocity Survey}
There are different options to consider in the design of a SN~Ia peculiar velocity survey, though at this point it is difficult to assess which is optimal
or most cost effective.  While there are a variety
of recently identified indicators of SN~Ia diversity that improve upon the 2-parameter model (SALT2) currently used to determine distances, 
there is not yet an umbrella model that simultaneously captures all indicators, their correlations, and characterizes the model residuals.
Constructing such a model is a parallel
endeavor that will benefit peculiar-velocity  and other SN~Ia science.  Instead of making specific recommendations, we make several general conclusions
based upon the above findings.

\begin{itemize}
\item Spectroscopic redshifts:  Peculiar velocities are extracted directly from redshift measurements.  Redshift uncertainties of $>0.5$\% contribute significantly
to the error budget.  We thus conclude that there is the need for spectroscopic $R>200$ host-galaxy redshifts.
\item Role of LSST:
LSST can be taken to be a SN~Ia discovery and distance machine, or a discovery machine only.   The nominal cadence of 
 produces sparse per-band light curves, which limits the intrinsic magnitude dispersion possible with LSST data alone.  Leaving the
 distance determination for other telescope, LSST should go for $\Omega$.
\item Follow-up resources.  21 mag.  2--4m-class telescopes.
\end{itemize}
\section{Conclusions}

>>>>

The precision in  $f\sigma_8$ derived from the baseline WFD ten-year SN~Ia discoveries has been projected by \citet{2017ApJ...847..128H},
from both their RSD and peculiar velocities.
The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance uncertainty is assumed for each supernova.
SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy, cluster, and Ly$\alpha$
RSD measurements are sensitive.  A joint galaxy RSD and SN peculiar velocity will be even more powerful.
SNe~Ia peculiar velocities will measure $f\sigma_8$ at $z<-.3$  was well as galaxy surveys will at $z \sim 0.6$ using RSD. 
{\bf Add a SN PV + DESI-like BGS column and/or a PV-only column with Cullan's code? }
% Requires the booktabs if the memoir class is not being used
\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{@{} lccc @{}} % Column formatting, @{} suppresses leading/trailing space
	\hline
	Redshift & RSD & RSD + PV & cumulative\\ \hline
      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\
      \hline
   \end{tabular}
   \caption{Projected percent uncertainties in $f\sigma_8$ from a 10-year LSST SN survey with 5\% distance uncertainties from
   \citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
   The cumulative column shows the effective uncertainty from combining the redshift and all other shallower redshift bins.}
   \label{tab:howlett}
\end{table}

@@

\section{Designing a Supernova Survey to Measure $\MakeLowercase{f}\sigma_8$}
\subsection{Survey Parameters that Determine Scientific Performance}
\citet{2017ApJ...847..128H} project how well peculiar velocity surveys can measure $f\sigma_8$. 
They perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the covariance for the velocity-velocity correlation
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\label{cov:eq}
\end{equation}
is dependent on the power spectrum, noise in the velocity measurement dominated by the intrinsic velocity dispersion $\sigma$, and the density of velocity probes
\citep{2017MNRAS.464.2517H}.  
The primary dependence on the growth of structure is through the normalization of the velocity power spectrum, such that 
\begin{equation}
 \frac{\partial P_{vv}}{\partial \lambda} = \text{constant}
\end{equation}
for the parameter choice
$\lambda=(f\sigma_8)^2$.  In the sample-variance limit, $F_{\lambda \lambda} \propto V$ whereas in the shot noise limit $F_{\lambda \lambda} \propto V n^2 \sigma^{-4}$.

The survey parameters that  enter the Fisher matrix are the volume $V$, which
in turn can be parameterized by the survey solid angle $\Omega$ and the redshift depth $z_{max}$ ; the number density of sources $n$; and the intrinsic
magnitude dispersion $\sigma_M$, where magnitude and velocity dispersions are related by $\sigma_{M} = \frac{5}{\ln{10}} \frac{1+z}{z} \sigma$.
LSST is expected to discover all $z<0.3$ SNe~Ia before maximum light in its active Wide Fast Deep (WFD) survey area.
Equating LSST discovery with having a meaningful peculiar velocity, the number density $n$ is thus a non-tunable parameter, which only increases with survey duration.
Thus $\Omega$, $z_{max}$, and $\sigma_M$ are left as the relevant parameters.

\begin{itemize}
\item Solid Angle $\Omega$:
The variance in $f\sigma_8$ is proportional to solid angle $\Omega$.  The baseline
LSST survey covers 18,000 sq.~deg, at $-75 \lesssim \delta \lesssim15$ avoiding the Galactic plane.  Larger solid-angle coverage for $z<0.3$ SN~Ia searches, beyond
the LSST baseline, benefits peculiar-velocity science.  Complementary northern-hemisphere surveys and LSST-expanded or independent
coverage of the southern equatorial pole, could double
the sky coverage and halve the variance in $f\sigma_8$.
\item Redshift Depth $z_{max}$:
For the number densities generated by LSST and the SN~Ia intrinsic magnitude dispersion, the noise is not sample-variance
limited.  We have run simulations of LSST SN surveys
and calculated the signal-to-noise (STON) of $f\sigma_8$ from the peculiar-velocity correlations. These simulations show, as seen in the left plot
of  Figure~\ref{notshot:fig}, that an increase in the number the number density of tracers leads to a significant improvement in STON.
It is also worth noting that sample variance is not negligible, as seen by the non-linear dependence of the  STON on $\sigma_M^2/n$.

@@

The effective contribution to the STON per supernova, quantified as STON divided by the root of the number of tracers, is higher for lower
$z_{max}$.  This is shown in the
right plot of   Figure~\ref{notshot:fig}.

A low-redshift tracer is more valuable than one at higher redshift.

\item Intrinsic Magnitude Dispersion $\sigma_M$:
A decrease in the supernova intrinsic magnitude dispersion leads to a significant improvement in  the STON of $f\sigma_8$,
as shown in Figure~\ref{notshot:fig} in parallel to our earlier discussion of the dependence of $n$.  We are not sample-variance
limited so the $\sigma^2_M/n$-term
in the Fisher matrix is important.
Survey follow-up stratey,
which determines which data are available per SN~Ia, influences the accuracy to which their distances can be determined. 
This dependence is non-trivial: the improvement between $\sigma_M=0.08$ and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or equivalently in survey duration.

The intrinsic magnitude dispersion depends on the purity of the sample. Conservatively we consider
a pure SN~Ia sample obtained through
spectroscopic classification.  The feasibility of using photometric classification is an important subject of research
that can have implications for survey planning.
\end{itemize}

\subsection{Designing an SN~Ia Peculiar Velocity Survey}
There are different options to consider in the design of a SN~Ia peculiar velocity survey, though at this point it is difficult to assess which is optimal
or most cost effective.  While there are a variety
of recently identified indicators of SN~Ia diversity that improve upon the 2-parameter model (SALT2) currently used to determine distances, 
there is not yet an umbrella model that simultaneously captures all indicators, their correlations, and characterizes the model residuals.
Here we  make several general conclusions
based upon the above findings, while noting that 
constructing such a model is a parallel
endeavor that will benefit peculiar-velocity  and other SN~Ia science.

\begin{itemize}
\item Spectroscopic transient classification:  SNe~Ia are defined through the absence of Hydrogen and the presence of the 6150~\AA~Silicon P-Cygni feature.
\item Spectroscopic redshifts:  Peculiar velocities are extracted directly from redshift measurements.  Redshift uncertainties of $>0.5$\% contribute significantly
to the error budget.  We thus conclude that there is the need for spectroscopic $R>200$ host-galaxy redshifts.  In most cases this redshift will be
available in the classification spectrum.  Planned bright-galaxy redshift surveys can also observe a significant fraction of nearby SN hosts.
\item Role of LSST:
LSST can be taken to be a SN~Ia discovery and distance machine, or a discovery machine only.   The nominal cadence of 
 produces sparse per-band light curves, which limits the intrinsic magnitude dispersion possible with LSST data alone to $\sigma_M \approx 0.15$~mag.
The wide-field of LSST is
 well suited for discoveries over a larger area of sky, but does not provide significant multiplex advantage for the low surface-density of active
$z<0.3$ SNe~Ia.  LSST is  suited for maximizing $\Omega$ while other (cheaper) resources can improve $\sigma_M$.
\item Follow-up resources to determine per-SN distances:
Supplemental non-LSST follow-up data in the form or improved temporal light-curve sampling (accurate rise and decline times), 
expanded (UV, NIR) wavelength coverage, and spectral features
can access high-fidelity SN~Ia models with lower intrinsic magnitude dispersion and residual systematic bias.
For example, infrared data \citep{2012MNRAS.425.1007B} or spectrophotometry at peak brightness
 \citep{2015ApJ...815...58F} are projected to give $\sigma_M \lesssim 0.08$~mag.

@@

A follow-up survey with $z_{max}=0.2$ would have $\sim 150,000$ targets over 10 years.  These sources will be observed when
brighter than $r \sim 20.5$~mag and so will be accessible to 2--4m-class telescopes.  
\end{itemize}

\section{Conclusions}
SNe~Ia are already powerful probes of the homogeneous cosmological expansion of the Universe.  In the next decade.
high-cadence, wide-field imaging surveys, together with improved precision in their distance determinations, will make SNe~Ia powerful probes
of the gravity induced-motion caused by the inhomogeneous Universe.  SNe~Ia peculiar velocities at $z<0.3$ will measure $f \sigma_8$ as well
as galaxy surveys, but at lower redshifts that provide better leverage to test gravity models.
While imaging surveys will provide a steady stream of SNe~Ia, a coordinated plan of follow-up is required to take advantage of
their probative power.  The resources necessary to follow hundreds of thousands of SNe depend on specific follow-up choices, and 
access to those resources will define the redshift limits of the survey.  Fortunately, the lowest-redshift, and hence brightest, supernovae
are of the highest interest, so that a modest suite of $\sim 5$ 2--4m telescopes should already provide leading measurements of
$f\sigma_8$ at $z\sim 0.1$.
